Hybrid quantum-classical unsupervised data clustering based on the Self-Organizing Feature Map

### 2020-09-22 manuscript.tex
REFACTOR to PNAS style

\documentclass[pra,showkeys,twocolumn,showpacs]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{array}
\usepackage{color}
\usepackage{soul}
%\usepackage{floatflt}
\usepackage{graphicx}
\DeclareMathOperator{\tr}{Tr}
\newcommand*{\eq}[1]{(\ref{#1})}
\newcommand*{\eqs}[2]{(\ref{#1})--(\ref{#2})}
\newcommand{\exi}{\tr{\{e^{-\frac\beta2\sum_k\lambda_k\gamma_k^\dag\gamma_k}\}}}
\renewcommand{\l}{\left(}
\renewcommand{\r}{\right)}

\usepackage[inline]{enumitem} % inline enumeration

\begin{document}
\title{Hybrid quantum-classical unsupervised data clustering based on the Self-Organizing Feature Map}

\author{I.~D.~Lazarev}
\affiliation{Institute of Problems of Chemical Physics of Russian Academy of Sciences, Acad. Semenov av. 1, Chernogolovka, Moscow Region, Russia, 142432}
\affiliation{ Faculty of Fundamental Physical-Chemical Engineering, Lomonosov Moscow State University, GSP-1, Moscow, Russia 119991}

\author{Marek Narozniak}
\affiliation{New York University Shanghai, 1555 Century Ave, Pudong, Shanghai 200122, China}
\affiliation{Department of Physics, New York University, New York, NY, 10003, USA.}

\author{Tim Byrnes}
\affiliation{New York University Shanghai, 1555 Century Ave, Pudong, Shanghai 200122, China}
\affiliation{State Key Laboratory of Precision Spectroscopy, School of Physical and Material Sciences, East China Normal University, Shanghai 200062, China}
\affiliation{NYU-ECNU Institute of Physics at NYU Shanghai, 3663 Zhongshan Road North, Shanghai 200062, China}
\affiliation{National Institute of Informatics, 2-1-2 Hitotsubashi, Chiyoda-ku, Tokyo 101-8430, Japan}
\affiliation{Department of Physics, New York University, New York, NY 10003, USA}

\author{A.~N.~Pyrkov}
\email{Email address:pyrkov@icp.ac.ru}
\affiliation{Institute of Problems of Chemical Physics of Russian Academy of Sciences, Acad. Semenov av. 1, Chernogolovka, Moscow Region, Russia, 142432}

\date{\today}

@@

%\pacs{}
\maketitle

\section{Introduction}
The combination of big data and artificial intelligence ---  dubbed the fourth industrial revolution --- has profoundly affected the modern economy in a plethora of different ways from robotics to agriculture \cite{Lecun2015, ghahramani2015,schwab2017,esteva2019, tyrsa2017}.
Contemporary artificial intelligence methods based on neural networks also have the potential to enhance the role of novel analytical methods in science and engineering \cite{kaggle2014, radovic2018, butler2018, radovic2018}.
Paradoxically, the intrinsic mechanism of how neural networks work and why they are so powerful remains unknown (in many cases it is regarded as a black box).
It has been speculated that limits to the neural network paradigm based on the computational power of von Neumann architecture are being approached,
and improvements appear only due to heuristic escalation of complexity \cite{marcus2018,sze2017,kourtis2020}.

@@

\section{The quantum assisted self-organizing feature map (QASOFM)}
\label{sec:qasofm}

\subsection{The classical algorithm}

@@

\subsection{Optimized quantum scheme for Hamming distance calculation}
\label{subsec:qcircuit}

@@

\section{Experimental demonstration of QASOFM}

@@

\section{Discussions}
We have developed a quantum assisted SOFM and showed a proof-of-concept experimental demonstration
that it can be used to solve clustering problems in an unsupervised manner.
The procedure of solving such a clustering problem requires calculating the distance many times in iterative way,
which is calculated using a hybrid quantum-classical procedure.
We introduced an optimized circuit for Hamming distance calculations that can be implemented on currently available quantum computing devices with high fidelity.
Our quantum circuit performs the distance-computing component of a classical SOFMs algorithm
and in this way improves its performance.
The complexity of the quantum assisted SOFM scales as $O(LN)$
while the complexity of the fully classical SOFM scales as $O(LMN)$,
where $N$ is number of samples, $M$ is number of randomly sampled cluster vectors,
and $L$ is number of the shifts of cluster vectors.
Due to wide use of classical SOFMs in different areas of modern research and technology,
this can give opportunities for the use of QASOFM in practical applications in the near term, outperforming classical algorithms.
In addition, as our algorithm performs the Hamming distance calculation,
it has the potential to enhance any classical algorithm that relies on calculating distances between data entries of vector form.
In machine learning, data science, statistics and optimization, distance is a common way of representing similarity, calculating it between large data sets is common procedure
and our circuit could potentially enhance other distance-based algorithms as long as exact distance is not required,
but when knowledge of nearest vectors is sufficient.

\section*{Acknowledgments}
The work is supported by the RFBR-NSFC collaborative program (Grant No. 18-57-53007) and the State assignment (N. AAAA-A19-119071190017-7). I.D.L. acknowledges support from the Russian Foundation for Basic Research, grant No.19-32-80004 and support from the Foundation for the Advancement of Theoretical Physics and Mathematics “BASIS” No.19-1-5-130-1. T.B. is supported by the Shanghai Research Challenge Fund; New York University Global Seed Grants for Collaborative Research; National Natural Science Foundation of China (61571301, D1210036A); the NSFC Research Fund for International Young Scientists (11650110425, 11850410426); NYU-ECNU Institute of Physics at NYU Shanghai; the Science and Technology Commission of Shanghai Municipality (17ZR1443600); the China Science and Technology Exchange Center (NGA-16-001); and the NSFC-RFBR Collaborative grant (81811530112).

@@

\section*{Conflict of interest}

The authors declare no conflict of interest.

\section*{Keywords}
quantum self-organizing feature map; quantum machine learning; neural network; quantum unsupervised data clustering; Hamming distance.

\bibliographystyle{naturemag}
\bibliography{bibliography}

>>>>

\PassOptionsToPackage{inline}{enumitem}
\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
\templatetype{pnasresearcharticle} % Choose template

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{array}
\usepackage{color}
\usepackage{soul}
%\usepackage{floatflt}
\usepackage{graphicx}

\DeclareMathOperator{\tr}{Tr}
\newcommand*{\eq}[1]{(\ref{#1})}
\newcommand*{\eqs}[2]{(\ref{#1})--(\ref{#2})}
\newcommand{\exi}{\tr{\{e^{-\frac\beta2\sum_k\lambda_k\gamma_k^\dag\gamma_k}\}}}
\renewcommand{\l}{\left(}
\renewcommand{\r}{\right)}

\title{Hybrid quantum-classical unsupervised data clustering based on the Self-Organizing Feature Map}

\author[a,b]{I.~D.~Lazarev}
\author[c,d]{Marek Narozniak}
\author[c,d,e,f,g]{Tim Byrnes}
\author[a,1]{A.~N.~Pyrkov}
\leadauthor{Pyrkov}

\affil[a]{Institute of Problems of Chemical Physics of Russian Academy of Sciences, Acad. Semenov av. 1, Chernogolovka, Moscow Region, Russia, 142432}
\affil[b]{ Faculty of Fundamental Physical-Chemical Engineering, Lomonosov Moscow State University, GSP-1, Moscow, Russia 119991}

\affil[c]{New York University Shanghai, 1555 Century Ave, Pudong, Shanghai 200122, China}
\affil[d]{Department of Physics, New York University, New York, NY, 10003, USA.}

\affil[e]{State Key Laboratory of Precision Spectroscopy, School of Physical and Material Sciences, East China Normal University, Shanghai 200062, China}
\affil[f]{NYU-ECNU Institute of Physics at NYU Shanghai, 3663 Zhongshan Road North, Shanghai 200062, China}
\affil[g]{National Institute of Informatics, 2-1-2 Hitotsubashi, Chiyoda-ku, Tokyo 101-8430, Japan}

%\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{The authors declare no conflict of interest.}

%\equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}E-mail: pyrkov\@icp.ac.ru}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.}

\keywords{
	quantum self-organizing feature map $|$
	quantum machine learning $|$
	neural network $|$
	quantum unsupervised data clustering $|$
	Hamming distance
}

@@

\date{\today}
%\doi{}

\begin{document}
\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\dropcap{T}he combination of big data and artificial intelligence ---  dubbed the fourth industrial revolution --- has profoundly affected the modern economy in a plethora of different ways from robotics to agriculture \cite{Lecun2015, ghahramani2015,schwab2017,esteva2019, tyrsa2017}.
Contemporary artificial intelligence methods based on neural networks also have the potential to enhance the role of novel analytical methods in science and engineering \cite{kaggle2014, radovic2018, butler2018, radovic2018}.
Paradoxically, the intrinsic mechanism of how neural networks work and why they are so powerful remains unknown (in many cases it is regarded as a black box).
It has been speculated that limits to the neural network paradigm based on the computational power of von Neumann architecture are being approached,
and improvements appear only due to heuristic escalation of complexity \cite{marcus2018,sze2017,kourtis2020}.

@@

\section*{The quantum assisted self-organizing feature map (QASOFM)}
\label{sec:qasofm}

\subsection*{The classical algorithm}

@@

\subsection*{Optimized quantum scheme for Hamming distance calculation}
\label{subsec:qcircuit}

@@

\section*{Experimental demonstration of QASOFM}

@@

\section*{Discussions}
We have developed a quantum assisted SOFM and showed a proof-of-concept experimental demonstration
that it can be used to solve clustering problems in an unsupervised manner.
The procedure of solving such a clustering problem requires calculating the distance many times in iterative way,
which is calculated using a hybrid quantum-classical procedure.
We introduced an optimized circuit for Hamming distance calculations that can be implemented on currently available quantum computing devices with high fidelity.
Our quantum circuit performs the distance-computing component of a classical SOFMs algorithm
and in this way improves its performance.
The complexity of the quantum assisted SOFM scales as $O(LN)$
while the complexity of the fully classical SOFM scales as $O(LMN)$,
where $N$ is number of samples, $M$ is number of randomly sampled cluster vectors,
and $L$ is number of the shifts of cluster vectors.
Due to wide use of classical SOFMs in different areas of modern research and technology,
this can give opportunities for the use of QASOFM in practical applications in the near term, outperforming classical algorithms.
In addition, as our algorithm performs the Hamming distance calculation,
it has the potential to enhance any classical algorithm that relies on calculating distances between data entries of vector form.
In machine learning, data science, statistics and optimization, distance is a common way of representing similarity, calculating it between large data sets is common procedure
and our circuit could potentially enhance other distance-based algorithms as long as exact distance is not required,
but when knowledge of nearest vectors is sufficient.

\acknow{
The work is supported by the RFBR-NSFC collaborative program (Grant No. 18-57-53007) and the State assignment (N. AAAA-A19-119071190017-7). I.D.L. acknowledges support from the Russian Foundation for Basic Research, grant No.19-32-80004 and support from the Foundation for the Advancement of Theoretical Physics and Mathematics “BASIS” No.19-1-5-130-1. T.B. is supported by the Shanghai Research Challenge Fund; New York University Global Seed Grants for Collaborative Research; National Natural Science Foundation of China (61571301, D1210036A); the NSFC Research Fund for International Young Scientists (11650110425, 11850410426); NYU-ECNU Institute of Physics at NYU Shanghai; the Science and Technology Commission of Shanghai Municipality (17ZR1443600); the China Science and Technology Exchange Center (NGA-16-001); and the NSFC-RFBR Collaborative grant (81811530112).
}
\showacknow{}

@@

%\bibliographystyle{naturemag}
\bibliography{bibliography}
